---
title: "Data Wrangling Note"
author: "Nneka Vivian Iduu"
date: "2025-03-19"
output: 
  html_document:
    keep_md: yes
  pdf_document: default
---

## Seamless DATA WRANGLING

We will need to install and load the tidyverse for easy handling of large datasets

### This note covers the DATA WRNAGLING and MANIPULATION functions which includes;
  -mutate
  -select
  -filter
  -the pipe %>%
  -summarize
  -group_by
  -joining
  -pivotting
  -Integration with plotting
  

```{r}
library(tidyverse)
```


The microbiome dataset, Bull_richness.csv dataset will be used for demonstrations
```{r}
microbiome.fungi <- read.csv("Bull_richness.csv")
str(microbiome.fungi)
```

## To select certain columns in dataset, we use the function select()

```{r}
microbiome.fungi2 <- select(microbiome.fungi, SampleID, Crop, Compartment:Fungicide, richness)
str(microbiome.fungi2)
```
Notice that column was used between Compartment:Fungicide,this helps to select columns that are in order. that is every column between the two columns including both of them

## To filter or subset data (i.e selecting some rows) using filter ()

### This function allows you to subset data fairly easily
```{r}
head(filter(microbiome.fungi2, Treatment == "Conv."))

# A more complex using &
head(filter(microbiome.fungi2, Treatment == "Conv." & Fungicide == "C"))

# Another more complex example using or |
head(filter(microbiome.fungi2, Sample == "A" | Sample == "B")) # samples A or B
```

## To create a new column using mutate()

```{r}
# Create a new column called logRich
head(mutate(microbiome.fungi2, logRich = log(richness))) # creates a new column logRich by taking the log of the values in column richness

# Creating a new column which combines Crop and Treatment 
head(mutate(microbiome.fungi2, Crop_Treatment = paste(Crop, Treatment)))
```

## To combine multiple functions together using pipe %>%

This is a new syntax that can allow us combine output of one into the input of another
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  head() # displaying the first six rows
```

## To summarize data using summarize()

This provides information like means and standard deviations/errors.
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich)) # calculating overall mean log richness within the conventionally managed treatment
```

In addition, if we want to have summary statistics by group, we can also do this very easily using the group_by() function. To calculate standard deviation and standard error

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```

## Connecting to plotting

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n)) %>%
  ggplot(aes(x = Fungicide, y = Mean.rich)) + # adding in a ggplot
  geom_bar(stat="identity") +
  geom_errorbar( aes(x=Fungicide, ymin=Mean.rich-std.err, ymax=Mean.rich+std.err), width=0.4) +
  theme_minimal() +
  xlab("") +
  ylab("Log Richness") +
  facet_wrap(~Treatment)
```


## To match up or join data using join(); more specifically 

  - left_join() - Keep all rows of X and add matching rows from Y. Any rows in Y that don’t match X are excluded.
  - right_join() - reverse of left_join() -inner_join() - only keep rows that are common to both X AND Y, remove everything else. 
  - full_join() - Keep any columns that are in either X or Y
  
```{r}
# selecting just the richness and sample ID
richness <- microbiome.fungi %>%
  select(SampleID, richness)

# selecting columns that don't include the richness
metadata <- microbiome.fungi %>% 
  select(SampleID, Fungicide, Crop, Compartment, GrowthStage, Treatment, Rep, Sample)

head(metadata)

head(richness)

head(left_join(metadata, richness, by = "SampleID")) # adding the richness data to the metadata based on on the common column of sampleID
```


## To convert from wide to long format and back again via pivoting with pivot_longer() and pivot_wider()

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) # calculates the mean per Treatment and Fungicide 

```

## Using  pivot_wider () which will set the values within the fungicide column into column names

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) # pivot to wide format
```

 We can easily can take the difference bewteen the fungicide and control now
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) %>% # pivot to wide format
  mutate(diff.fungicide = C - F) # calculate the difference between the means. 
```

Now we can easily calculate the difference between fungicide and control and plot it.
```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) %>% # pivot to wide format
  mutate(diff.fungicide = C - F) %>%  # calculate the difference between the means. 
  ggplot(aes(x = Treatment, y = diff.fungicide)) + # Plot it 
  geom_col() +
  theme_minimal() +
  xlab("") +
  ylab("Difference in average species richness")
```

